---
title: "A Spatial Analysis of Walkability and Urban Form in the Chicago Metropolitan Area"
author:
  - name: Shanmukh Upadhyayula
    affiliation: Department of Geography and GIS, University of Illinois at Urbana-Champaign
date: "December 18, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
editor_options:
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-family: "Times New Roman", Times, serif;
}

h1, h2, h3, h4, h5, h6 {
  font-family: "Times New Roman", Times, serif;
}

pre, code {
  font-family: "Courier New", Courier, monospace;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r setwd, include=FALSE}
setwd("C:/Users/shanm/Documents/Personal Projects/ShanmukhUpad.github.io")
dir.create("poster_figures", showWarnings = FALSE)
```

```{r environment-setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(tmap)
library(knitr)
library(corrplot)
library(broom)
library(spdep)
library(scales)
library(tigris)

options(tigris_use_cache = TRUE)
```

# **Introduction**

Walkability is a measure for how safe an area is for pedestrians as well
as amenities being within a reasonable walking distance. Right now, the
ChiVes database does not have a walkability measure for the built
environment of Chicago. Thus, adding such a measure would give an idea
of what the City of Chicago could do to improve the built environment.
Chicago itself already has a decently high walkability score compared to
most cities in the United States, but some areas still suffer from
car-friendly exclusive infrastructure. Improving the walkability score
for a region would also improve the urban sustainability and the
transportation equity of the city.

The data source for this project is the EPA National Walkability Index
from 2021. This dataset uses block groups as its spatial scale, but to
assure compatibility with ChiVes, it has been converted to the 2020
census tracts.

**Hypothesis:** Walkability will be higher in the CBD/The Loop compared
to outer areas. Since there is a high population density near the
center, walkability should also high. However, although it is expected
that most census tracts will have an above average walkability score due
to Chicago being the third most populous city in the United States,
there will still be a decent chunk of census tracts in the below average
range as well as the most walkable range. Thus, this project will also
include population density and transit access as exploratory variables.

The goal this project is to visualize data so that there is a way to see
how walkabilty varies across Chicago and whether or not higher transit
access areas have a direct, inverse, or unrelated correlation with
walkability. Such visualizations already exist out in on the internet,
but adding it to ChiVes will place it among many other important
environmental data so that everyone can see this measure in one coherent
place.

------------------------------------------------------------------------

# **Data Wrangling**

The dataset used for this project is the EPA National Walkability
geodatabase from 2021. The data has been filtered to only include the
Chicago metropolitan area by filtering the CBSA and the FIPS codes to
the region. Next, the data must be cleaned to remove any possible
analysis to ensure accuracy of the data. Therefore, anything with
missing values have been removed.

The way that the EPA classifies walkability is out of a score of 20 that
goes like the following:

**EPA Walkability Score Categories:**\
1 - 5.75: Least Walkable\
5.76 - 10.50: Below Average Walkable\
10.51 - 15.25: Above Average Walkable\
15.26 - 20: Most Walkable

The main problem with this is that the "most walkable" category is too
large of a range, even if it uses the same difference of 4.76/4.75 per
category. Since this is out of 20, a census tract with a score of 15.26
would be in the 76th percentile, while a census tract even just 3 points
higher would be in the 91st percentile. That is why it is important to
add another category. 15.26 + 4.76 = 20.02, which would make this fifth
category above the scale itself. That's why for this project, the fifth
category, most walkable (the category previously called most walkable is
now renamed to highly walkable), is now only at 18.1, or just slightly
above the 90th percentile.

**Adjusted Categories:**

15.26 - 18.09: Highly Walkable

18.10 - 20: Most Walkable

The spatial scale that this data is in the format of is that of block
groups. However, since ChiVes uses census tracts, the data is converted
from block groups to census tracts. Since a census tracts compose of
many different block groups, it is expected that there should be more
block groups than census tracts. If it is the opposite, then something
would be done incorrectly in the calculations. Thus, it makes sense why
the number of block groups is 2.9 times higher than the number of census
tracts.

Once all the cleaning is done, it is finally time to create categorical
walkability variables. The EPA rates walkability on a scale from 0-20.
In the code, the ranges on how walkability is categorized are shown in
the **r transformations** section. The sum of all walkability category
variables should equal the number of census tracts after cleaning the
data.

```{r load-data}
# Load the National Walkability Index data from geodatabase
walkability_raw <- st_read("WalkabilityIndex/Natl_WI.gdb",
                           layer = "NationalWalkabilityIndex")
cat("Variables available:", ncol(walkability_raw), "\n\n")
names(walkability_raw)
```

```{r spatial-filter}
# Filter to Chicago metropolitan area
walkability_chicago <- walkability_raw %>%
  filter(STATEFP == "17", CBSA == "16980") %>%
  st_transform(crs = 4326)

cat("Number of block groups in Chicago metropolitan area:", nrow(walkability_chicago), "\n")
```

```{r variable-selection}
# Extract census tract ID and select variables
walkability_clean <- walkability_chicago %>%
  mutate(TRACT_GEOID = substr(GEOID10, 1, 11)) %>%
  select(
    GEOID_BG = GEOID20,
    TRACT_GEOID,
    walk_score = NatWalkInd,
    employment_entropy = D2A_EPHHM,
    employment_mix     = D2B_E8MIXA,
    intersection_density = D3B,
    transit_distance = D4A,
    total_pop = TotPop,
    total_workers = Workers,
    households = HH,
    land_area_ac = Ac_Land,
    geometry = Shape
  ) %>%
  # EPA encodes missing transit data as -99999; remove before aggregation
  mutate(transit_distance = ifelse(transit_distance < 0, NA, transit_distance))

cat("Block groups before aggregation:", nrow(walkability_clean), "\n")
```

```{r data-cleaning}
# Check for missing values
missing_summary <- walkability_clean %>%
  st_drop_geometry() %>%
  summarise(
    missing_walk_score = sum(is.na(walk_score)),
    missing_pop = sum(is.na(total_pop)),
    missing_transit = sum(is.na(transit_distance)),
    missing_land_area = sum(is.na(land_area_ac)),
    total_block_groups = n()
  )

print(missing_summary)

# Remove block groups with missing data
walkability_clean <- walkability_clean %>%
  filter(!is.na(walk_score), !is.na(total_pop), !is.na(land_area_ac),
         land_area_ac > 0, walk_score > 0)

cat("\nBlock groups after cleaning:", nrow(walkability_clean), "\n")
```

```{r block-groups-to-census-tracts}
# Aggregate from block groups to census tracts
# Population-weighted mean with per-variable NA handling. weighted.mean()'s
# na.rm option drops NA elements from x without dropping the matching weights,
# which trips its own length check; this helper does it correctly.
pw_mean <- function(x, w) {
  ok <- !is.na(x) & !is.na(w) & is.finite(x) & is.finite(w)
  total_w <- sum(w[ok])
  if (length(ok) == 0 || !any(ok) || !is.finite(total_w) || total_w == 0)
    return(NA_real_)
  sum(x[ok] * w[ok]) / total_w
}

tract_data <- walkability_clean %>%
  filter(!is.na(walk_score), !is.na(total_pop), !is.na(land_area_ac)) %>%
  group_by(TRACT_GEOID) %>%
  summarise(
    # NatWalkInd is a linear combination of D2A_EPHHM, D2B_E8MIXA, D3B, and
    # D4A at the block-group level. To preserve that linear identity at the
    # tract level, every input is aggregated population-weighted with
    # per-variable NA handling, so that a block group missing one input,
    # for example the EPA sentinel for transit distance, does not get
    # dropped from the aggregation of the other inputs. The earlier
    # behavior, where any sentinel forced the entire block group out of
    # every aggregation, was what was suppressing the bivariate r values.
    # Order matters: every pw_mean call has to consume the original
    # block-group-level total_pop vector. If total_pop is summarised first,
    # subsequent pw_mean calls receive a scalar weight, which collapses ok-
    # indexing in pw_mean to NA and turns most tract values into NA.
    walk_score           = pw_mean(walk_score,           total_pop),
    intersection_density = pw_mean(intersection_density, total_pop),
    employment_entropy   = pw_mean(employment_entropy,   total_pop),
    employment_mix       = pw_mean(employment_mix,       total_pop),
    transit_distance     = pw_mean(transit_distance,     total_pop),
    total_workers        = sum(total_workers, na.rm = TRUE),
    households           = sum(households,    na.rm = TRUE),
    land_area_ac         = sum(land_area_ac),
    total_pop            = sum(total_pop),
    .groups = 'drop'
  ) %>%
  mutate(
    pop_density_sqmi = (total_pop / land_area_ac) * 640,
    workers_per_hh = total_workers / (households + 0.1),
    transit_distance_km = transit_distance / 1000
  )

cat("\nNumber of census tracts:", nrow(tract_data), "\n")
```

```{r load-chives}
# Load ChiVes data
chives_data <- st_read("chives-data-public.geojson")

chives_clean <- chives_data %>%
  st_drop_geometry() %>%
  mutate(geoid = as.character(geoid))

tract_data_clean <- tract_data %>%
  mutate(TRACT_GEOID = as.character(TRACT_GEOID))

tract_data_with_chives <- tract_data_clean %>%
  left_join(chives_clean, by = c("TRACT_GEOID" = "geoid"))
```

```{r city-tracts}
# Census tracts in city of Chicago
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city_tracts <- tract_data_with_chives %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

num_chicago_tracts <- nrow(chicago_city_tracts)
difference <- nrow(tract_data_with_chives) - num_chicago_tracts

# Print results
cat("Census tracts in the city of Chicago:", num_chicago_tracts, "\n")
cat("Difference from Greater Chicago area census tracts:", difference, "\n")
cat("Percentage of census tracts in the city of Chicago:",
    floor((num_chicago_tracts / nrow(tract_data_with_chives)) * 1000) / 10, "%\n")
```

```{r categorization}
# Create categorical walkability variable with adjusted categories
final_data <- tract_data_with_chives %>%
  mutate(
    walk_category = case_when(
      walk_score >= 18.1 ~ "Most Walkable",
      walk_score >= 15.26 & walk_score < 18.1 ~ "Highly Walkable",
      walk_score >= 10.51 & walk_score < 15.26 ~ "Above Average Walkable",
      walk_score >= 5.76 & walk_score < 10.51 ~ "Below Average Walkable",
      walk_score >= 1 & walk_score < 5.76 ~ "Least Walkable",
      TRUE ~ "No Data"
    ),
    walk_category = factor(
      walk_category,
      levels = c(
        "Least Walkable",
        "Below Average Walkable",
        "Above Average Walkable",
        "Highly Walkable",
        "Most Walkable",
        "No Data"
      )
    )
  )

# city_data is the canonical analysis frame for hypothesis-scope statistics.
# The hypothesis is about the City of Chicago (CBD vs outer neighborhoods),
# so all bivariate correlations, the corrplot, the scatter, the scatter
# matrix, the linear regression, and the ANOVA are computed on this subset.
# Maps of metro-wide spatial pattern still use the full final_data.
city_data <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects)
```

------------------------------------------------------------------------

# **Exploratory Methods**

The main approach for this model was to create statistical variables
such as finding the lowest, highest, mean, median, and standard
deviation walk scores. In addition, the mean population density and mean
transit distance has been calculated in order to support the main
walkability variables.

Once the variables are created, correlations were found among them via a
correlation matrix. Then, a few statistical tests were created around
these variables such as a Moran I's test, linear regression, and ANOVA.
These exploratory methods allow spatial analysis that provide important
insights into Chicago.

**Reasons for each method's inclusion:**\

**Moran I's Test:**

This test measures spatial autocorrelations, which are important in
finding out whether census tracts with high or low walkability scores
cluster together. For Chicago, it can show if the most walkable
neighborhoods like The Loop connect easily with surrounding
neighborhoods, or if the walkability is contained within the
neighborhood itself.

**Linear Regression and Correlation Matrix:**

These statistical analysis quantify the relationship between the main
walkability variable and exploratory variables such as population
density, transit access, street intersection density, and employment
entropy. Seeing the relationships between walkability and these
variables can influence the sustainability and overall qaulity of life
in Chicago.

**Analysis of Variance (ANOVA):**

This method compares the mean differences between contiuous variables
across all five levels of walkability. Thus, it is important for seeing
whether there is much of a difference in these variables between a level
like "Below Average Walkable" vs "Highly Walkable".

```{r descriptive-stats}
# Summary statistics
summary_stats <- final_data %>%
  st_drop_geometry() %>%
  summarise(
    N_Tracts = n(),
    Mean_Walkability = round(mean(walk_score, na.rm = TRUE), 2),
    SD_Walkability = round(sd(walk_score, na.rm = TRUE), 2),
    Median_Walk = round(median(walk_score, na.rm = TRUE), 2),
    Min_Walk = round(min(walk_score, na.rm = TRUE), 2),
    Max_Walk = round(max(walk_score, na.rm = TRUE), 2),
    Mean_Pop_Density = round(mean(pop_density_sqmi, na.rm = TRUE), 0),
    Mean_Transit_Dist = round(mean(transit_distance_km, na.rm = TRUE), 2)
  )

kable(summary_stats, caption = "Summary Statistics for Chicago Census Tracts")
```

```{r correlation}
# Correlation matrix on the City of Chicago analysis frame.
# Predictors with negligible bivariate correlation with walkability at city
# extent (|r| < 0.20) are excluded; pop_density_sqmi (r = -0.05) is dropped.
cor_data <- city_data %>%
  st_drop_geometry() %>%
  select(walk_score, intersection_density, employment_entropy,
         employment_mix, transit_distance_km) %>%
  na.omit()

cor_matrix <- cor(cor_data, use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix: Walkability and Related Variables (City of Chicago)",
         mar = c(0,0,2,0))
```

```{r spatial-autocorrelation}
# Moran's I test for spatial autocorrelation
final_data_complete <- final_data %>%
  filter(!is.na(walk_score))

coords <- st_coordinates(st_centroid(final_data_complete))
neighbors <- knn2nb(knearneigh(coords, k = 8))
weights <- nb2listw(neighbors, style = "W")

moran_test <- moran.test(final_data_complete$walk_score, weights)
cat("Moran's I test for spatial autocorrelation:\n")
print(moran_test)

if(moran_test$p.value < 0.05) {
  cat("\nSignificant spatial autocorrelation detected (p < 0.05)")
} else {
  cat("\nNo significant spatial autocorrelation detected")
}
```

```{r statistical-tests}
# Linear regression on the City of Chicago analysis frame
model <- lm(walk_score ~ pop_density_sqmi + transit_distance_km +
              intersection_density + employment_entropy + employment_mix,
            data = city_data)

summary(model)

model_summary <- tidy(model)
kable(model_summary, digits = 3,
      caption = "Regression Results: Predictors of Walkability Score (City of Chicago)")

# ANOVA on the City of Chicago analysis frame
anova_result <- aov(pop_density_sqmi ~ walk_category, data = city_data)
print(summary(anova_result))

anova_pvalue <- summary(anova_result)[[1]]$`Pr(>F)`[1]
if(!is.na(anova_pvalue) && anova_pvalue < 0.05) {
  cat("\nTukey HSD Post-hoc test:\n")
  print(TukeyHSD(anova_result))
}
```

------------------------------------------------------------------------

# **Results**

The data reflects the initial hypothesis that most areas would be in
the "Above Average Walkability" range. 1,090 out of 2,014 census tracts
are in this range. The hypothesis also predicted "a decent chunk" of
census tracts in the most walkable range; however, only 20 out of 2,014
(1%) fall into the "Most Walkable" category (≥ 18.10). An additional
350 (17.4%) are "Highly Walkable," so the prediction holds better when
interpreted as the top two tiers combined.

The spatial half of the hypothesis, that walkability is highest in the
CBD/Loop and falls off in outer neighborhoods, is supported by the
maps and by the Moran's I test, which shows significant positive
spatial autocorrelation. The urban-form half of the hypothesis is also
supported once the EPA inputs are aggregated to tract level the same
way the walkability score is (population-weighted). Within the City of
Chicago, employment + household entropy (D2A_EPHHM) correlates
strongly with walkability (r ≈ 0.69), the 8-tier employment mix
(D2B_E8MIXA) is comparably strong (r ≈ 0.68), intersection density
(D3B) is also strong (r ≈ 0.53), and transit distance (D4A) is weak
negative (r ≈ −0.25). These are the underlying components of the EPA formula,
which is why side-by-side choropleth maps look so visually similar:
central neighborhoods light up across all four inputs simultaneously,
and the Pearson r values now match that visual impression. The earlier
weak r values were an aggregation artifact: block groups missing one
input were being dropped from every aggregation, distorting the tract
values for the inputs they did have.

The one piece of the hypothesis the data does *not* support is
population density acting as the driver: city-extent r is essentially
zero. Density is not an EPA input, and within Chicago the densest
tracts (Lakeview, Edgewater) are not the most walkable; the Loop is,
despite a comparatively low residential density. The mechanism the
data points to is therefore mixed-use development, dense
intersections, and transit proximity, not raw residential density.

**Walkability Map for the City of Chicago versus Walkability Map for the
Greater Chicago Area:\
**The visualization for the walkability scores of each census tract in
Chicago describes what the paragraph above has mentioned. The larger the
census tract's land area, the more likely its walkability score is that
of a lower percentile. Since the ChiVes website only visualizes the city
of Chicago itself, only the main walkability map for the city will be
used, but including the Greater Chicago area shows how suburban sprawl
has affected the walkability of America, as many other cities have
similar patterns to Chicago.

**Higher Variability Walkability Map for the City of Chicago:**

Although the adjusted fourth category and the addition of a fifth
category added more variability among census tracts to produce a
visualization at a higher accuracy, using a jenks style provides natural
breaks in the data, which can produce a visualization where there is an
even higher variability among census tracts. This uses its own
categorization, which does not even count for the "Least Walkable"
category, as not a single census tract within the city itself is
categorized as such.

**Street Intersection Density:**

This map helps in identifying pedestrian-centric or car-centric areas,
which is helpful for understanding the impacts of a built environment's
impact on walkability and sustainability. The density of a street
intersection may reveal its proximity to transit stations or other
amenities that a city has to offer. A jenks classification style has
been used for this map, as it uses natural breaks in the data. For a
street network, it minimizes and class variances to show groupings more
easily than quantile would.

**Employment + Household Entropy Scatter Plot:**

This scatter plot shows the relationship between walkability and
D2A_EPHHM (employment + household entropy), one of the four EPA
inputs to the walkability index. Within the City of Chicago this is
the strongest bivariate predictor of walkability (r ≈ 0.69,
R² ≈ 0.48): tracts with greater mixed-use intensity score
substantially higher on walkability. The strong, near-linear positive
trend is exactly what we should expect because D2A_EPHHM enters the
EPA's NatWalkInd formula directly.

**Economic Opportunity Visualization:**

This map has been created using the economic hardship variable taken
from the existing ChiVes dataset. The opportunity variable is the
inverse of hardship. The purpose of this visualization's inclusion in
this analysis is to reveal correlations between walkability and a
positive variable such as economic opportunity. Census tracts with the
highest economic opportunity tend, on average, to fall into the higher
walkability categories; the bivariate correlation within the City of
Chicago is r ≈ 0.37 (moderate), so individual tracts still deviate
from the pattern. A jenks classification style has been used for this
map, for the same reasons as the street intersection density map.

**Distance to Nearest Transit Stop Visualization:**

This map shows the population-weighted distance from each tract to its
nearest transit stop. Within the city the bivariate correlation with
walkability is r ≈ −0.25 (weak, negative): tracts farther from
transit tend to score lower. Transit distance is one of the four
direct inputs to the EPA walkability formula, but its city-level
range is compressed because most Chicago tracts sit close to a
transit stop, which suppresses the bivariate r relative to the other
EPA inputs.

**LISA Cluster Maps (Local Moran's I):**

Two LISA maps are included: one for the Greater Chicago metropolitan
area and one for the City of Chicago alone, with the city-only
neighbor weights recomputed inside the city boundary so that spatial
lag does not bleed across the city limit. Each map shows the
significant (p < 0.05) local clusters: High-High tracts (high
walkability surrounded by high-walkability neighbors), Low-Low
tracts, and the High-Low and Low-High spatial outliers. Both maps
are gated on a significant global Moran's I; if the global test were
not significant, the local quadrants would not be interpretable and
the map is suppressed.

**Scatterplot Matrix:**

This visualization shows pairwise relationships between walkability and
the urban-form variables retained for the City of Chicago analysis.
Population density is excluded because its city-extent correlation
with walkability is negligible (r ≈ −0.05); the remaining panels
reflect weak-to-strong associations.

```{r results-summary}
# Summary table by walkability category
results_table <- final_data %>%
  st_drop_geometry() %>%
  group_by(walk_category) %>%
  summarise(
    N_Tracts = n(),
    Mean_Walk_Score = round(mean(walk_score, na.rm = TRUE), 2),
    Total_Population = sum(total_pop, na.rm = TRUE),
    Mean_Pop_Density = round(mean(pop_density_sqmi, na.rm = TRUE), 0),
    Mean_Transit_Dist_km = round(mean(transit_distance_km, na.rm = TRUE), 2),
    Mean_Intersection_Density = round(mean(intersection_density, na.rm = TRUE), 1)
  )

kable(results_table, 
      caption = "Mean Values by Walkability Category (Census Tract Level)")
```

```{r greater-chicago-walkability, fig.width=10, fig.height=8}
metro_walk_map <- tm_shape(final_data) +
  tm_fill(
    "walk_category",
    palette = c(
      "Least Walkable"         = "#440154",
      "Below Average Walkable" = "#31688E",
      "Above Average Walkable" = "#35B779",
      "Highly Walkable"        = "#90D743",
      "Most Walkable"          = "#FDE725",
      "No Data"                = "#bdbdbd"
    ),
    title = "Walkability Category"
  ) +
  tm_borders(alpha = 0.3, lwd = 0.5) +
  tm_layout(
    main.title = "Greater Chicago Area Walkability by Census Tract (2021)",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.outside.position = "right",
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(position = c("left", "bottom"))
metro_walk_map
```

```{r chicago-city-walkability, fig.width=10, fig.height=8}
# Get Chicago city boundary from Census
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

# Main walkability score map
city_walk_map <- tm_shape(chicago_city) +
  tm_fill(
    "walk_category",
    palette = c(
      "Least Walkable"         = "#440154",
      "Below Average Walkable" = "#31688E",
      "Above Average Walkable" = "#35B779",
      "Highly Walkable"        = "#90D743",
      "Most Walkable"          = "#FDE725",
      "No Data"                = "#bdbdbd"
    ),
    title = "Walkability Category"
  ) +
  tm_borders(alpha = 0.3, lwd = 0.5) +
  tm_layout(
    main.title = "Chicago City Walkability by Census Tract (2021)",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.outside.position = "right",
    frame = TRUE
  ) +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(position = c("left", "bottom"))
city_walk_map
```

```{r variability-visualizations, fig.width=10, fig.height=8}

# Walkability score map with higher variations
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

tm0 <- tm_shape(chicago_city) +
  tm_fill("walk_score",
          palette = "viridis",
          style = "jenks",
          n = 5,
          title = "Walkability Category") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Chicago City Walkability Scores Adjusted for Higher Standards of Chicago",
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)
tmap_arrange(tm0, ncol = 1)
```

```{r additional-visualizations, fig.width=12, fig.height=6}
# Chicago city boundary
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

# Maps restricted to variables with at least a moderate bivariate correlation
# with walkability inside the City of Chicago. Population density (r ≈ -0.05)
# and workers per household (r ≈ -0.01) are excluded as negligible.

# Visualization 1: Street Intersection Density (city r ≈ 0.53, strong)
tm2 <- tm_shape(chicago_city) +
  tm_fill("intersection_density",
          palette = "Oranges",
          style = "jenks",
          n = 5,
          title = "Intersections/km²") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Street Intersection Density",
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)

# Visualization 2: Economic Opportunity (city r ≈ 0.37, moderate)
chicago_city_quality <- chicago_city %>%
  filter(!is.na(hrdshpIndx)) %>%
  mutate(opportunity = 100 - hrdshpIndx)

tm3 <- tm_shape(chicago_city_quality) +
  tm_fill("opportunity",
          palette = "-magma",
          style = "jenks",
          n = 5,
          title = "Opportunity Index") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Economic Opportunity (Lower Hardship)",
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)

# Visualization 3: Transit Distance (city r ≈ -0.25, weak negative)
tm5 <- tm_shape(chicago_city) +
  tm_fill("transit_distance_km",
          palette = "-BuPu",
          style = "jenks",
          n = 5,
          title = "Distance (km)") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Distance to Nearest Transit Stop",
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)

tmap_arrange(tm2, tm3, tm5, ncol = 3)
```

```{r lisa-metro, fig.width=10, fig.height=8}
# Local Indicators of Spatial Association across the Greater Chicago metro.
# Only run if the global Moran's I detected significant clustering, since
# LISA quadrants are uninterpretable without underlying spatial autocorrelation.
if (moran_test$p.value < 0.05) {
  local_m <- localmoran(final_data_complete$walk_score, weights)
  lag_walk <- lag.listw(weights, final_data_complete$walk_score)

  walk_z <- as.numeric(scale(final_data_complete$walk_score))
  lag_z  <- as.numeric(scale(lag_walk))
  p_lisa <- local_m[, 5]

  quadrant <- rep("Not Significant", length(walk_z))
  sig <- !is.na(p_lisa) & p_lisa < 0.05
  quadrant[sig & walk_z >  0 & lag_z >  0] <- "High-High"
  quadrant[sig & walk_z <  0 & lag_z <  0] <- "Low-Low"
  quadrant[sig & walk_z >  0 & lag_z <  0] <- "High-Low"
  quadrant[sig & walk_z <  0 & lag_z >  0] <- "Low-High"

  final_data_complete <- final_data_complete %>%
    mutate(lisa_cluster = factor(
      quadrant,
      levels = c("High-High", "Low-Low", "High-Low", "Low-High", "Not Significant")
    ))

  cat("LISA cluster counts (Greater Chicago, p < 0.05):\n")
  print(table(final_data_complete$lisa_cluster))

  print(
    tm_shape(final_data_complete) +
      tm_fill(
        "lisa_cluster",
        palette = c(
          "High-High"       = "#b2182b",
          "Low-Low"         = "#2166ac",
          "High-Low"        = "#f4a582",
          "Low-High"        = "#92c5de",
          "Not Significant" = "#f0f0f0"
        ),
        title = "LISA cluster (p < 0.05)"
      ) +
      tm_borders(alpha = 0.2, lwd = 0.3) +
      tm_layout(
        main.title = "Local Moran's I Clusters of Walkability (Greater Chicago)",
        main.title.size = 1.1,
        legend.outside = TRUE,
        legend.outside.position = "right",
        frame = FALSE
      )
  )
} else {
  cat("Skipping LISA: global Moran's I was not significant, so local clusters",
      "would not be interpretable.\n")
}
```

```{r lisa-city, fig.width=10, fig.height=8}
# City-only LISA. Recompute neighbors and weights inside the City of Chicago
# so the spatial lag does not bleed across the city boundary, then run a
# city-only Moran's I as a gate before the local cluster map.
city_lisa_data <- city_data %>% filter(!is.na(walk_score))

city_coords    <- st_coordinates(st_centroid(city_lisa_data))
city_neighbors <- knn2nb(knearneigh(city_coords, k = 8))
city_weights   <- nb2listw(city_neighbors, style = "W")

city_moran <- moran.test(city_lisa_data$walk_score, city_weights)
cat("Moran's I (City of Chicago):\n"); print(city_moran)

if (city_moran$p.value < 0.05) {
  local_m_city <- localmoran(city_lisa_data$walk_score, city_weights)
  lag_walk_city <- lag.listw(city_weights, city_lisa_data$walk_score)

  walk_z_c <- as.numeric(scale(city_lisa_data$walk_score))
  lag_z_c  <- as.numeric(scale(lag_walk_city))
  p_lisa_c <- local_m_city[, 5]

  quadrant_c <- rep("Not Significant", length(walk_z_c))
  sig_c <- !is.na(p_lisa_c) & p_lisa_c < 0.05
  quadrant_c[sig_c & walk_z_c >  0 & lag_z_c >  0] <- "High-High"
  quadrant_c[sig_c & walk_z_c <  0 & lag_z_c <  0] <- "Low-Low"
  quadrant_c[sig_c & walk_z_c >  0 & lag_z_c <  0] <- "High-Low"
  quadrant_c[sig_c & walk_z_c <  0 & lag_z_c >  0] <- "Low-High"

  city_lisa_data <- city_lisa_data %>%
    mutate(lisa_cluster = factor(
      quadrant_c,
      levels = c("High-High", "Low-Low", "High-Low", "Low-High", "Not Significant")
    ))

  cat("LISA cluster counts (City of Chicago, p < 0.05):\n")
  print(table(city_lisa_data$lisa_cluster))

  city_lisa_map <- tm_shape(city_lisa_data) +
    tm_fill(
      "lisa_cluster",
      palette = c(
        "High-High"       = "#b2182b",
        "Low-Low"         = "#2166ac",
        "High-Low"        = "#f4a582",
        "Low-High"        = "#92c5de",
        "Not Significant" = "#f0f0f0"
      ),
      title = "LISA cluster (p < 0.05)"
    ) +
    tm_borders(alpha = 0.2, lwd = 0.3) +
    tm_layout(
      main.title = "Local Moran's I Clusters of Walkability (City of Chicago)",
      main.title.size = 1.1,
      legend.outside = TRUE,
      legend.outside.position = "right",
      frame = TRUE
    )
  print(city_lisa_map)
} else {
  cat("Skipping city LISA: city-only Moran's I was not significant.\n")
}
```

```{r walkability-correlations}
# Bivariate correlations within the City of Chicago.
# Variables with negligible city-extent r are excluded (pop_density_sqmi, r = -0.05).
city_stats <- city_data %>%
  st_drop_geometry() %>%
  select(walk_score, intersection_density, employment_entropy,
         employment_mix, transit_distance_km) %>%
  na.omit()

city_quality_stats <- chicago_city_quality %>%
  st_drop_geometry() %>%
  select(walk_score, opportunity) %>%
  na.omit()

ct_int <- cor.test(city_stats$walk_score, city_stats$intersection_density)
ct_emp <- cor.test(city_stats$walk_score, city_stats$employment_entropy)
ct_mix <- cor.test(city_stats$walk_score, city_stats$employment_mix)
ct_trn <- cor.test(city_stats$walk_score, city_stats$transit_distance_km)
ct_opp <- cor.test(city_quality_stats$walk_score, city_quality_stats$opportunity)

tibble(
  Variable  = c("Street Intersection Density",
                "Employment + HH Entropy (D2A)",
                "8-tier Employment Mix (D2B)",
                "Transit Distance (km)",
                "Economic Opportunity"),
  r         = round(c(ct_int$estimate, ct_emp$estimate, ct_mix$estimate,
                      ct_trn$estimate, ct_opp$estimate), 3),
  `R²`      = round(r^2, 3),
  `p-value` = c(ct_int$p.value, ct_emp$p.value, ct_mix$p.value,
                ct_trn$p.value, ct_opp$p.value),
  Sig       = dplyr::case_when(
    `p-value` < 0.001 ~ "***",
    `p-value` < 0.01  ~ "**",
    `p-value` < 0.05  ~ "*",
    TRUE              ~ "n.s."
  )
) %>%
  mutate(`p-value` = ifelse(`p-value` < 0.001, "< 0.001",
                            as.character(round(`p-value`, 3)))) %>%
  knitr::kable(caption = "Bivariate Pearson correlations with walkability score within the City of Chicago, after population-weighted aggregation of all EPA inputs (matching the aggregation used for walk_score). Population density and workers per household are excluded as negligible (|r| < 0.1).")
```

```{r regression-visualization, fig.width=10, fig.height=6, message=FALSE}
# Scatter for the strongest bivariate predictor: D2A employment + household
# entropy (one of the four EPA inputs to NatWalkInd).
scatter_data <- city_data %>% st_drop_geometry() %>%
  select(walk_score, employment_entropy) %>% na.omit()

r_emp  <- round(cor(scatter_data$walk_score, scatter_data$employment_entropy), 3)
r2_emp <- round(r_emp^2, 3)

ggplot(scatter_data, aes(x = employment_entropy, y = walk_score)) +
  geom_point(alpha = 0.35, color = "#2c7fb8") +
  geom_smooth(method = "lm", se = TRUE, color = "#253494") +
  scale_y_continuous(limits = c(0, 20)) +
  labs(
    title = "Employment + Household Entropy vs Walkability (City of Chicago)",
    subtitle = paste0("r = ", r_emp, ", R² = ", r2_emp,
                      ". D2A_EPHHM is one of the four EPA inputs to the walkability index"),
    x = "Employment + Household Entropy (D2A_EPHHM)",
    y = "Walkability Score",
    caption = "Each point = one Chicago city census tract"
  ) +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11, color = "#555"))
```

```{r scatterplot-matrix, fig.width=8, fig.height=6}
# Scatterplot matrix on City of Chicago analysis frame
pairs_data <- city_data %>%
  st_drop_geometry() %>%
  select(walk_score, transit_distance_km,
         intersection_density, employment_entropy, employment_mix) %>%
  na.omit()

pairs(pairs_data,
      main = "Scatterplot Matrix: Walkability and Related Variables (City of Chicago)",
      pch = 16,
      col = rgb(0, 0, 1, 0.3))
```

------------------------------------------------------------------------

# **Discussion**

This analysis was scoped to the City of Chicago because that is the
extent the original hypothesis was about: walkability higher in the
CBD/Loop and falling off in outer neighborhoods. Three findings
support the hypothesis. First, the Moran's I test confirms that
walkability is not randomly distributed but follows clear
neighborhood-scale spatial patterns, and the maps show walkability is
highest in and around the Loop, exactly as predicted. Second, once
the EPA inputs are aggregated to tract level the same way the
walkability score is, population-weighted with per-variable NA
handling so that block groups missing one input are not dropped from
every aggregation, the bivariate correlations within Chicago are
strong: employment + household entropy (D2A_EPHHM) reaches r ≈ 0.69,
the 8-tier employment mix (D2B_E8MIXA) reaches r ≈ 0.68, intersection
density (D3B) reaches r ≈ 0.53, and transit distance (D4A) reaches
r ≈ −0.25. These are the underlying components of the
EPA walkability formula and they correlate with the score by
construction, which is exactly what makes the side-by-side
choropleth maps look so visually similar. Third, the multivariate
regression on Chicago city tracts explains R² ≈ 0.85 of walkability
variance, which is what we should expect because the predictors are
the formula inputs.

The one piece of the hypothesis the data does not support is the idea
that raw residential population density drives walkability. Within
the city, Pearson r between pop_density_sqmi and walkability is
essentially zero. The reason is mechanistic, not statistical: density
is not one of the four EPA inputs, and within Chicago the densest
residential tracts (Lakeview, Edgewater, parts of Lincoln Park) are
not the most walkable tracts. The Loop has the highest walkability
but a comparatively low residential density because it is dominated
by jobs and retail. The variables that do drive walkability are
mixed-use development, dense street networks, and transit proximity,
all of which the Loop has in abundance.

Across the United States, an emerging experimental trend is the
"15-minute city", where all the amenities that come with being in a
town or city can be accessed within the span of a 15-minute walk.
Even in some low-density suburbs, there is demand for a reduction in
car-centric infrastructure. The Chicago data is consistent with this
framing: a tract can be relatively low-density and still score high
on walkability if mixed-use entropy and intersection density are
high. The strongest individual driver in the multivariate regression
is the 8-tier employment mix, which is exactly the variable that
captures whether jobs, retail, services, and housing are colocated at
walkable distances within a tract.

All of these findings justify the inclusion of this dataset into ChiVes
as it provides this information within the context of all of the other
environmental factors that either positively or negatively influence the
lives of Chicagoans on a daily basis. However, there are still a few
limitations to this dataset that prevented this analysis from reaching
its full potential. There are five census tracts with no data, due to
them primarily not having many people at all. Furthermore, walkability
scores do not account for a few aspects of the urban form, such as the
quality of a sidewalk, how convenient it is to walk somewhere in warmer
versus colder weather, crime rates, or how relevant an amenity is on a
daily basis to the average person.

Throughout the analysis, several issues with the R code occurred. The
first challenge was converting from block groups to census tracts. The
initial approach to this was taking a weighted mean of the block groups,
but that would have produced misleading results due to the variables for
the data being counted incorrectly (i.e. not counting some data or
double counting some data). It is also challenging to include a census
tract if it is missing some variables. Thus, removing all data for the
census tract is the easiest thing to do. Thankfully, this only removed
5/2014 census tracts. This method of computing weighted means was later
changed to sum the product of a variable's value and its corresponding
weight, divided by the sum of the weights. In other words, this method
accounted for the population of an area, since census tracts with larger
populations contributed more proportionally to the final conversion from
block groups to census tracts than lower density areas.

The next challenge was getting the exact Chicago boundaries. Initially,
the spatial filtering to visualize Chicago was just using four
coordinates to form a rectangle around the area. The problem with this
was that areas outside the desired analysis were present in the
visualization. Thus, the spatial filtering was later changed to using
places function from the tigris library and transforming the location
into the correct CRS. This filtered the boundaries to the exact polygon.

Then, another persistent challenge to overcome in this analysis was
surprisingly loading in the ChiVes data. Usually, loading in a dataset
should be simple, but the EPA Walkability variables were being
incorporated into the chicago_city final data before the ChiVes
variables had been incorporated into that same variable. This was solved
by performing a left_join() on the EPA Walkability's TRACT_GEOID and
ChiVes' geoid variable.

This analysis is only an introductory step in a hopefully ongoing
process to bring attention to improve walkability across not just
Chicago but around all of the Greater Chicago Metropolitan area as a
whole. Implementing this into ChiVes would vastly increase the chances
of new policy implementation such as new transit lines, added bike
lanes, re-zoning projects, and new street layout projects. As previously
mentioned four paragraphs back, a walkability score does not account for
everything related to the convenience of pedestrians. Future work for
this project could include adding data to integrate these factors into
new visualizations. Perhaps an entirely new data generating process
(DGP) is needed in order to maximize the accuracy of this data. In fact,
the idea that 1090 census tracts are classified as "Above Average
Walkability" does not make sense. If the average walkability score is
singular value, it makes sense why there is no category called "Average
Walkability"; Everything can only either be below or above that average.
Finally, training machine learning models can speed up and also increase
the accuracy of the aforementioned DGP. These models would be able to
predict walkability scores based off of current built-environment
variables, create new variables, and even incorporate scenario-based
testing.

All of these goals and this current analysis is extremely important for
the environmental sustainability and quality of life in the urban form.

------------------------------------------------------------------------

# **References and .csv Export**

```{r export-to-csv}
# Restrict export to the exact tract set in ChiVes so the join in the ChiVes
# database is 1:1 with no extra tracts.
chives_geoids <- chives_clean %>% pull(geoid) %>% unique()

chicago_export <- final_data %>%
  st_drop_geometry() %>%
  filter(TRACT_GEOID %in% chives_geoids) %>%
  select(
    geoid = TRACT_GEOID,
    walkability_score = walk_score,
    walkability_category = walk_category,
    employment_entropy = employment_entropy,
    intersection_density = intersection_density,
    transit_distance_km = transit_distance_km,
  )

cat("Tracts exported to ChiVes:", nrow(chicago_export), "\n")
cat("ChiVes tracts total:", length(chives_geoids), "\n")
cat("ChiVes tracts missing walkability:",
    length(setdiff(chives_geoids, chicago_export$geoid)), "\n")

write_csv(chicago_export, "chicago_walkability_for_chives.csv")

```

-   U.S. Environmental Protection Agency. (2021). National Walkability
    Index: User Guide and Methodology.
-   U.S. Environmental Protection Agency. (2021). Smart Location
    Database Technical Documentation.
-   ChiVes Project Team. (2024). ChiVes: Chicago Visualization and
    Environmental Sustainability Platform. University of Illinois at
    Urbana-Champaign.

------------------------------------------------------------------------
