---
title: "A Spatial Analysis of Walkability and Urban Form in the Chicago Metropolitan Area"
author:
  - name: Shanmukh Upadhyayula
    affiliation: Department of Geography and GIS, University of Illinois at Urbana-Champaign
date: "December 18, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
editor_options:
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-family: "Times New Roman", Times, serif;
}

h1, h2, h3, h4, h5, h6 {
  font-family: "Times New Roman", Times, serif;
}

pre, code {
  font-family: "Courier New", Courier, monospace;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r setwd, include=FALSE}
setwd("C:/Users/shanm/Documents/Personal Projects/ShanmukhUpad.github.io")
```

```{r environment-setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(tmap)
library(knitr)
library(corrplot)
library(broom)
library(spdep)
library(scales)
library(tigris)

options(tigris_use_cache = TRUE)
```

# **Introduction**

Walkability is a measure for how safe an area is for pedestrians as well
as amenities being within a reasonable walking distance. Right now, the
ChiVes database does not have a walkability measure for the built
environment of Chicago. Thus, adding such a measure would give an idea
of what the City of Chicago could do to improve the built environment.
Chicago itself already has a decently high walkability score compared to
most cities in the United States, but some areas still suffer from
car-friendly exclusive infrastructure. Improving the walkability score
for a region would also improve the urban sustainability and the
transportation equity of the city.

The data source for this project is the EPA National Walkability Index
from 2021. This dataset uses block groups as its spatial scale, but to
assure compatibility with ChiVes, it has been converted to the 2020
census tracts.

**Hypothesis:** Walkability will be higher in the CBD/The Loop compared
to outer areas. Since there is a high population density near the
center, walkability should also high. However, although it is expected
that most census tracts will have an above average walkability score due
to Chicago being the third most populous city in the United States,
there will still be a decent chunk of census tracts in the below average
range as well as the most walkable range. Thus, this project will also
include population density and transit access as exploratory variables.

The goal this project is to visualize data so that there is a way to see
how walkabilty varies across Chicago and whether or not higher transit
access areas have a direct, inverse, or unrelated correlation with
walkability. Such visualizations already exist out in on the internet,
but adding it to ChiVes will place it among many other important
environmental data so that everyone can see this measure in one coherent
place.

------------------------------------------------------------------------

# **Data Wrangling**

The dataset used for this project is the EPA National Walkability
geodatabase from 2021. The data has been filtered to only include the
Chicago metropolitan area by filtering the CBSA and the FIPS codes to
the region. Next, the data must be cleaned to remove any possible
analysis to ensure accuracy of the data. Therefore, anything with
missing values have been removed.

The way that the EPA classifies walkability is out of a score of 20 that
goes like the following:

**EPA Walkability Score Categories:**\
1 - 5.75: Least Walkable\
5.76 - 10.50: Below Average Walkable\
10.51 - 15.25: Above Average Walkable\
15.26 - 20: Most Walkable

The main problem with this is that the "most walkable" category is too
large of a range, even if it uses the same difference of 4.76/4.75 per
category. Since this is out of 20, a census tract with a score of 15.26
would be in the 76th percentile, while a census tract even just 3 points
higher would be in the 91st percentile. That is why it is important to
add another category. 15.26 + 4.76 = 20.02, which would make this fifth
category above the scale itself. That's why for this project, the fifth
category, most walkable (the category previously called most walkable is
now renamed to highly walkable), is now only at 18.1, or just slightly
above the 90th percentile.

**Adjusted Categories:**

15.26 - 18.09: Highly Walkable

18.10 - 20: Most Walkable

The spatial scale that this data is in the format of is that of block
groups. However, since ChiVes uses census tracts, the data is converted
from block groups to census tracts. Since a census tracts compose of
many different block groups, it is expected that there should be more
block groups than census tracts. If it is the opposite, then something
would be done incorrectly in the calculations. Thus, it makes sense why
the number of block groups is 2.9 times higher than the number of census
tracts.

Once all the cleaning is done, it is finally time to create categorical
walkability variables. The EPA rates walkability on a scale from 0-20.
In the code, the ranges on how walkability is categorized are shown in
the **r transformations** section. The sum of all walkability category
variables should equal the number of census tracts after cleaning the
data.

```{r load-data}
# Load the National Walkability Index data from geodatabase
walkability_raw <- st_read("WalkabilityIndex/Natl_WI.gdb",
                           layer = "NationalWalkabilityIndex")
cat("Variables available:", ncol(walkability_raw), "\n\n")
names(walkability_raw)
```

```{r spatial-filter}
# Filter to Chicago metropolitan area
walkability_chicago <- walkability_raw %>%
  filter(STATEFP == "17", CBSA == "16980") %>%
  st_transform(crs = 4326)

cat("Number of block groups in Chicago metropolitan area:", nrow(walkability_chicago), "\n")
```

```{r variable-selection}
# Extract census tract ID and select variables
walkability_clean <- walkability_chicago %>%
  mutate(TRACT_GEOID = substr(GEOID10, 1, 11)) %>%
  select(
    GEOID_BG = GEOID20,
    TRACT_GEOID,
    walk_score = NatWalkInd,
    employment_entropy = D2A_EPHHM,
    intersection_density = D3B,
    transit_distance = D4A,
    total_pop = TotPop,
    total_workers = Workers,
    households = HH,
    land_area_ac = Ac_Land,
    geometry = Shape
  ) %>%
  # EPA encodes missing transit data as -99999; remove before aggregation
  mutate(transit_distance = ifelse(transit_distance < 0, NA, transit_distance))

cat("Block groups before aggregation:", nrow(walkability_clean), "\n")
```

```{r data-cleaning}
# Check for missing values
missing_summary <- walkability_clean %>%
  st_drop_geometry() %>%
  summarise(
    missing_walk_score = sum(is.na(walk_score)),
    missing_pop = sum(is.na(total_pop)),
    missing_transit = sum(is.na(transit_distance)),
    missing_land_area = sum(is.na(land_area_ac)),
    total_block_groups = n()
  )

print(missing_summary)

# Remove block groups with missing data
walkability_clean <- walkability_clean %>%
  filter(!is.na(walk_score), !is.na(total_pop), !is.na(land_area_ac),
         land_area_ac > 0, walk_score > 0)

cat("\nBlock groups after cleaning:", nrow(walkability_clean), "\n")
```

```{r block-groups-to-census-tracts}
# Aggregate from block groups to census tracts
tract_data <- walkability_clean %>%
  filter(!is.na(walk_score), !is.na(total_pop), !is.na(land_area_ac),
         !is.na(intersection_density), !is.na(employment_entropy),
         !is.na(transit_distance)) %>%
  group_by(TRACT_GEOID) %>%
  summarise(
    walk_score = sum(walk_score * total_pop) / sum(total_pop),
    total_pop = sum(total_pop),
    total_workers = sum(total_workers),
    households = sum(households),
    land_area_ac = sum(land_area_ac),
    intersection_density = sum(intersection_density * land_area_ac) / sum(land_area_ac),
    employment_entropy = sum(employment_entropy * total_pop) / sum(total_pop),
    transit_distance = sum(transit_distance * total_pop) / sum(total_pop),
    .groups = 'drop'
  ) %>%
  mutate(
    pop_density_sqmi = (total_pop / land_area_ac) * 640,
    workers_per_hh = total_workers / (households + 0.1),
    transit_distance_km = transit_distance / 1000
  )

cat("\nNumber of census tracts:", nrow(tract_data), "\n")
```

```{r load-chives}
# Load ChiVes data
chives_data <- st_read("chives-data-public.geojson")

chives_clean <- chives_data %>%
  st_drop_geometry() %>%
  mutate(geoid = as.character(geoid))

tract_data_clean <- tract_data %>%
  mutate(TRACT_GEOID = as.character(TRACT_GEOID))

tract_data_with_chives <- tract_data_clean %>%
  left_join(chives_clean, by = c("TRACT_GEOID" = "geoid"))
```

```{r city-tracts}
# Census tracts in city of Chicago
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city_tracts <- tract_data_with_chives %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

num_chicago_tracts <- nrow(chicago_city_tracts)
difference <- nrow(tract_data_with_chives) - num_chicago_tracts

# Print results
cat("Census tracts in the city of Chicago:", num_chicago_tracts, "\n")
cat("Difference from Greater Chicago area census tracts:", difference, "\n")
cat("Percentage of census tracts in the city of Chicago:", 
    floor((num_chicago_tracts / nrow(tract_data_with_chives)) * 1000) / 10, "%\n")
```

```{r categorization}
# Create categorical walkability variable with adjusted categories
final_data <- tract_data_with_chives %>%
  mutate(
    walk_category = case_when(
      walk_score >= 18.1 ~ "Most Walkable",
      walk_score >= 15.26 & walk_score < 18.1 ~ "Highly Walkable",
      walk_score >= 10.51 & walk_score < 15.26 ~ "Above Average Walkable",
      walk_score >= 5.76 & walk_score < 10.51 ~ "Below Average Walkable",
      walk_score >= 1 & walk_score < 5.76 ~ "Least Walkable",
      TRUE ~ "No Data"
    ),
    walk_category = factor(
      walk_category,
      levels = c(
        "Least Walkable",
        "Below Average Walkable",
        "Above Average Walkable",
        "Highly Walkable",
        "Most Walkable",
        "No Data"
      )
    )
  )
```

------------------------------------------------------------------------

# **Exploratory Methods**

The main approach for this model was to create statistical variables
such as finding the lowest, highest, mean, median, and standard
deviation walk scores. In addition, the mean population density and mean
transit distance has been calculated in order to support the main
walkability variables.

Once the variables are created, correlations were found among them via a
correlation matrix. Then, a few statistical tests were created around
these variables such as a Moran I's test, linear regression, and ANOVA.
These exploratory methods allow spatial analysis that provide important
insights into Chicago.

**Reasons for each method's inclusion:**\

**Moran I's Test:**

This test measures spatial autocorrelations, which are important in
finding out whether census tracts with high or low walkability scores
cluster together. For Chicago, it can show if the most walkable
neighborhoods like The Loop connect easily with surrounding
neighborhoods, or if the walkability is contained within the
neighborhood itself.

**Linear Regression and Correlation Matrix:**

These statistical analysis quantify the relationship between the main
walkability variable and exploratory variables such as population
density, transit access, street intersection density, and employment
entropy. Seeing the relationships between walkability and these
variables can influence the sustainability and overall qaulity of life
in Chicago.

**Analysis of Variance (ANOVA):**

This method compares the mean differences between contiuous variables
across all five levels of walkability. Thus, it is important for seeing
whether there is much of a difference in these variables between a level
like "Below Average Walkable" vs "Highly Walkable".

```{r descriptive-stats}
# Summary statistics
summary_stats <- final_data %>%
  st_drop_geometry() %>%
  summarise(
    N_Tracts = n(),
    Mean_Walkability = round(mean(walk_score, na.rm = TRUE), 2),
    SD_Walkability = round(sd(walk_score, na.rm = TRUE), 2),
    Median_Walk = round(median(walk_score, na.rm = TRUE), 2),
    Min_Walk = round(min(walk_score, na.rm = TRUE), 2),
    Max_Walk = round(max(walk_score, na.rm = TRUE), 2),
    Mean_Pop_Density = round(mean(pop_density_sqmi, na.rm = TRUE), 0),
    Mean_Transit_Dist = round(mean(transit_distance_km, na.rm = TRUE), 2)
  )

kable(summary_stats, caption = "Summary Statistics for Chicago Census Tracts")
```

```{r correlation}
# Correlation matrix - restricted to predictors with at least modest |r| with walkability
cor_data <- final_data %>%
  st_drop_geometry() %>%
  select(walk_score, intersection_density,
         employment_entropy, transit_distance_km) %>%
  na.omit()

cor_matrix <- cor(cor_data, use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix: Walkability and Related Variables",
         mar = c(0,0,2,0))
```

```{r spatial-autocorrelation}
# Moran's I test for spatial autocorrelation
final_data_complete <- final_data %>%
  filter(!is.na(walk_score))

coords <- st_coordinates(st_centroid(final_data_complete))
neighbors <- knn2nb(knearneigh(coords, k = 8))
weights <- nb2listw(neighbors, style = "W")

moran_test <- moran.test(final_data_complete$walk_score, weights)
cat("Moran's I test for spatial autocorrelation:\n")
print(moran_test)

if(moran_test$p.value < 0.05) {
  cat("\nSignificant spatial autocorrelation detected (p < 0.05)")
} else {
  cat("\nNo significant spatial autocorrelation detected")
}
```

```{r statistical-tests}
# Linear regression
model <- lm(walk_score ~ pop_density_sqmi + transit_distance_km + 
              intersection_density + employment_entropy, 
            data = final_data)

summary(model)

model_summary <- tidy(model)
kable(model_summary, digits = 3, 
      caption = "Regression Results: Predictors of Walkability Score")

# ANOVA
anova_result <- aov(pop_density_sqmi ~ walk_category, data = final_data)
print(summary(anova_result))

anova_pvalue <- summary(anova_result)[[1]]$`Pr(>F)`[1]
if(!is.na(anova_pvalue) && anova_pvalue < 0.05) {
  cat("\nTukey HSD Post-hoc test:\n")
  print(TukeyHSD(anova_result))
}
```

------------------------------------------------------------------------

# **Results**

The data reflects the initial hypothesis that most areas would be in the
"Above Average Walkability" range. 1,090 out of 2,014 census tracts are
in this range. The hypothesis also predicted "a decent chunk" of census
tracts in the most walkable range; however, only 20 out of 2,014 (1%)
fall into the "Most Walkable" category (≥ 18.10). An additional 350
(17.4%) are "Highly Walkable," so the prediction holds better when
interpreted as the top two tiers combined. The hypothesis that
population density would be a primary driver of walkability is *not*
supported by the bivariate data: Pearson r between walk_score and
pop_density_sqmi is only 0.114 (R² = 0.013) over the metro, which is a
weak association. Population density was therefore dropped from the
chart panels in favor of variables that show at least a modest
correlation with walkability (intersection density, economic
opportunity, transit distance).

**Walkability Map for the City of Chicago versus Walkability Map for the
Greater Chicago Area:\
**The visualization for the walkability scores of each census tract in
Chicago describes what the paragraph above has mentioned. The larger the
census tract's land area, the more likely its walkability score is that
of a lower percentile. Since the ChiVes website only visualizes the city
of Chicago itself, only the main walkability map for the city will be
used, but including the Greater Chicago area shows how suburban sprawl
has affected the walkability of America, as many other cities have
similar patterns to Chicago.

**Higher Variability Walkability Map for the City of Chicago:**

Although the adjusted fourth category and the addition of a fifth
category added more variability among census tracts to produce a
visualization at a higher accuracy, using a jenks style provides natural
breaks in the data, which can produce a visualization where there is an
even higher variability among census tracts. This uses its own
categorization, which does not even count for the "Least Walkable"
category, as not a single census tract within the city itself is
categorized as such.

**Street Intersection Density:**

This map helps in identifying pedestrian-centric or car-centric areas,
which is helpful for understanding the impacts of a built environment's
impact on walkability and sustainability. The density of a street
intersection may reveal its proximity to transit stations or other
amenities that a city has to offer. A jenks classification style has
been used for this map, as it uses natural breaks in the data. For a
street network, it minimizes and class variances to show groupings more
easily than quantile would.

**Economic Opportunity Scatter Plot:**

This scatter plot shows the relationship between walkability and
economic opportunity (the inverse of the ChiVes hardship index) for
each Chicago city census tract that has hardship data. Economic
opportunity is the strongest *bivariate* predictor of walkability in
this dataset (r ≈ 0.37, R² ≈ 0.14). The relationship is positive but
modest: tracts with higher opportunity tend to have higher walkability
scores, but a large amount of tract-level variation remains
unexplained.

**Economic Opportunity Visualization:**

This map has been created using the economic hardship variable taken
from the existing ChiVes dataset. The opportunity variable is the
inverse of hardship. The purpose of this visualization's inclusion in
this analysis is to reveal correlations between walkability and a
positive variable such as economic opportunity. Census tracts with the
highest economic opportunity tend, on average, to fall into the higher
walkability categories, but the bivariate correlation (r ≈ 0.37) is
only modest, so individual tracts deviate from the pattern. A jenks
classification style has been used for this map, for the same reasons
as the street intersection density map.

**Scatterplot Matrix:**

This visualization shows pairwise relationships between walkability and
the variables retained after pruning weak correlations. Variables with
|r| < 0.25 against walkability (population density, workers per
household) have been excluded so that every panel reflects at least a
modest association.

```{r results-summary}
# Summary table by walkability category
results_table <- final_data %>%
  st_drop_geometry() %>%
  group_by(walk_category) %>%
  summarise(
    N_Tracts = n(),
    Mean_Walk_Score = round(mean(walk_score, na.rm = TRUE), 2),
    Total_Population = sum(total_pop, na.rm = TRUE),
    Mean_Pop_Density = round(mean(pop_density_sqmi, na.rm = TRUE), 0),
    Mean_Transit_Dist_km = round(mean(transit_distance_km, na.rm = TRUE), 2),
    Mean_Intersection_Density = round(mean(intersection_density, na.rm = TRUE), 1)
  )

kable(results_table, 
      caption = "Mean Values by Walkability Category (Census Tract Level)")
```

```{r greater-chicago-walkability, fig.width=10, fig.height=8}
tm_shape(final_data) +
  tm_fill(
    "walk_category",
    palette = c(
      "Least Walkable" = "#ca0020",
      "Below Average Walkable" = "#f4a582",
      "Above Average Walkable" = "#f7f7f7",
      "Highly Walkable" = "#92c5de",
      "Most Walkable" = "#0571b0",
      "No Data" = "#bdbdbd"
    ),
    title = "Walkability Category"
  ) +
  tm_borders(alpha = 0.3, lwd = 0.5) +
  tm_layout(
    main.title = "Greater Chicago Area Walkability by Census Tract (2021)",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.outside.position = "right",
    frame = FALSE
  ) +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(position = c("left", "bottom"))
```

```{r chicago-city-walkability, fig.width=10, fig.height=8}
# Get Chicago city boundary from Census
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

# Main walkability score map
tm_shape(chicago_city) +
  tm_fill(
    "walk_category",
    palette = c(
      "Least Walkable" = "#ca0020",
      "Below Average Walkable" = "#f4a582",
      "Above Average Walkable" = "#f7f7f7",
      "Highly Walkable" = "#92c5de",
      "Most Walkable" = "#0571b0",
      "No Data" = "#bdbdbd"
    ),
    title = "Walkability Category"
  ) +
  tm_borders(alpha = 0.3, lwd = 0.5) +
  tm_layout(
    main.title = "Chicago City Walkability by Census Tract (2021)",
    main.title.size = 1.2,
    legend.outside = TRUE,
    legend.outside.position = "right",
    frame = TRUE
  ) +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(position = c("left", "bottom"))
```

```{r variability-visualizations, fig.width=10, fig.height=8}

# Walkability score map with higher variations
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

tm0 <- tm_shape(chicago_city) +
  tm_fill("walk_score",
          palette = "magma", 
          style = "jenks",
          n = 5,
          title = "Walkability Category") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Chicago City Walkability Scores Adjusted for Higher Standards of Chicago", 
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)
tmap_arrange(tm0, ncol = 1)
```

```{r additional-visualizations, fig.width=12, fig.height=6}
# Chicago city boundary
chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_city <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects)

# Maps restricted to variables with at least modest correlation with walkability.
# Population density (r=0.114) and workers per household (r=-0.079) were dropped
# because their bivariate correlation with walkability is too weak to interpret
# spatial co-variation visually.

# Visualization 1: Street Intersection Density
tm2 <- tm_shape(chicago_city) +
  tm_fill("intersection_density",
          palette = "Oranges",
          style = "jenks",
          n = 5,
          title = "Intersections/km²") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Street Intersection Density",
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)

# Visualization 2: Economic Opportunity (Inverse of Hardship)
chicago_city_quality <- chicago_city %>%
  filter(!is.na(hrdshpIndx)) %>%
  mutate(opportunity = 100 - hrdshpIndx)

tm3 <- tm_shape(chicago_city_quality) +
  tm_fill("opportunity",
          palette = "Purples",
          style = "jenks",
          n = 5,
          title = "Opportunity Index") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Economic Opportunity (Lower Hardship)",
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)

# Visualization 3: Transit Distance
tm5 <- tm_shape(chicago_city) +
  tm_fill("transit_distance_km",
          palette = "-BuPu",
          style = "jenks",
          n = 5,
          title = "Distance (km)") +
  tm_borders(alpha = 0.2, lwd = 0.3) +
  tm_layout(main.title = "Distance to Nearest Transit Stop",
            legend.position = c("left", "bottom"),
            main.title.size = 1.1,
            frame = TRUE)

tmap_arrange(tm2, tm3, tm5, ncol = 3)
```

```{r walkability-correlations}
# Bivariate correlations: walkability vs each mapped variable
# Restricted to variables with at least modest |r| (>= 0.25)
city_stats <- final_data %>%
  st_drop_geometry() %>%
  select(walk_score, intersection_density, transit_distance_km) %>%
  na.omit()

city_quality_stats <- chicago_city_quality %>%
  st_drop_geometry() %>%
  select(walk_score, opportunity) %>%
  na.omit()

ct_int <- cor.test(city_stats$walk_score, city_stats$intersection_density)
ct_opp <- cor.test(city_quality_stats$walk_score, city_quality_stats$opportunity)
ct_trn <- cor.test(city_stats$walk_score, city_stats$transit_distance_km)

tibble(
  Variable  = c("Street Intersection Density",
                "Economic Opportunity",
                "Transit Distance (km)"),
  r         = round(c(ct_int$estimate, ct_opp$estimate, ct_trn$estimate), 3),
  `R²`      = round(r^2, 3),
  `p-value` = c(ct_int$p.value, ct_opp$p.value, ct_trn$p.value),
  Sig       = dplyr::case_when(
    `p-value` < 0.001 ~ "***",
    `p-value` < 0.01  ~ "**",
    `p-value` < 0.05  ~ "*",
    TRUE              ~ "n.s."
  )
) %>%
  mutate(`p-value` = ifelse(`p-value` < 0.001, "< 0.001",
                            as.character(round(`p-value`, 3)))) %>%
  knitr::kable(caption = "Bivariate Pearson correlations with walkability score (two-tailed). Variables with |r| < 0.25 (population density, workers per household) excluded.")
```

```{r regression-visualization, fig.width=10, fig.height=6, message=FALSE}
# Scatter for the strongest bivariate predictor (Economic Opportunity, r ~ 0.37)
scatter_data <- chicago_city_quality %>% st_drop_geometry() %>%
  select(walk_score, opportunity) %>% na.omit()

r2_opp <- round(cor(scatter_data$walk_score, scatter_data$opportunity)^2, 3)
r_opp  <- round(cor(scatter_data$walk_score, scatter_data$opportunity), 3)

ggplot(scatter_data, aes(x = opportunity, y = walk_score)) +
  geom_point(alpha = 0.35, color = "#6a51a3") +
  geom_smooth(method = "lm", se = TRUE, color = "#3f007d") +
  scale_y_continuous(limits = c(0, 20)) +
  labs(
    title = "Economic Opportunity vs Walkability (City of Chicago)",
    subtitle = paste0("r = ", r_opp, ", R² = ", r2_opp,
                      " — strongest bivariate predictor among modest-correlated variables"),
    x = "Economic Opportunity Index (100 - Hardship)",
    y = "Walkability Score",
    caption = "Each point = one Chicago city census tract with hardship data"
  ) +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11, color = "#555"))
```

```{r scatterplot-matrix, fig.width=8, fig.height=6}
# Scatterplot matrix - restricted to predictors with at least modest |r| with walkability
pairs_data <- final_data %>%
  st_drop_geometry() %>%
  select(walk_score, transit_distance_km,
         intersection_density, employment_entropy) %>%
  na.omit()

pairs(pairs_data,
      main = "Scatterplot Matrix: Walkability and Related Variables",
      pch = 16,
      col = rgb(0, 0, 1, 0.3))
```

------------------------------------------------------------------------

# **Discussion**

This analysis reveals that walkability in the Chicago metropolitan area
has only a *weak* bivariate correlation with population density
(r = 0.114, R² = 0.013), so the initial hypothesis that high population
density is a direct driver of higher walkability is **not** supported by
the data, even though the multivariate regression still finds a
significant negative coefficient for pop_density_sqmi after controlling
for transit distance, intersection density, and employment entropy.
The Moran's I test confirms that walkability is not randomly
distributed but follows patterns on a neighborhood scale. This can be
seen in the walkability map, as walkability scores generally lower as
one leaves the city, with several exceptions.

Across the United States, an emerging experimental trend is the
"15-minute city", where all the amenities that come with being in a town
or city can be accessed within the span of a 15-minute walk. Even in
some low-density suburbs, there is demand for a reduction in car-centric
infrastructure. Among the variables examined here, only three show at
least a modest bivariate correlation with the walkability score:
intersection density (r = 0.355), economic opportunity (r = 0.369), and
transit distance (r = −0.276). Population density (r = 0.114) and
workers per household (r = −0.079) were dropped from the chart panels
because their associations are weak, even where the p-values are
statistically significant given the large sample size. Where amenities
are closer together, intersection density rises and mixed-use
buildings become more common, which is consistent with the modest
positive correlation observed between intersection density and
walkability. There are still many census tracts that deviate from this
general pattern — the multivariate regression explains R² ≈ 0.43 of
walkability variance, leaving most variation unaccounted for, so an
area can still be relatively low-density and have a high walkability
score, which is what the "15-minute city" aims for.

All of these findings justify the inclusion of this dataset into ChiVes
as it provides this information within the context of all of the other
environmental factors that either positively or negatively influence the
lives of Chicagoans on a daily basis. However, there are still a few
limitations to this dataset that prevented this analysis from reaching
its full potential. There are five census tracts with no data, due to
them primarily not having many people at all. Furthermore, walkability
scores do not account for a few aspects of the urban form, such as the
quality of a sidewalk, how convenient it is to walk somewhere in warmer
versus colder weather, crime rates, or how relevant an amenity is on a
daily basis to the average person.

Throughout the analysis, several issues with the R code occurred. The
first challenge was converting from block groups to census tracts. The
initial approach to this was taking a weighted mean of the block groups,
but that would have produced misleading results due to the variables for
the data being counted incorrectly (i.e. not counting some data or
double counting some data). It is also challenging to include a census
tract if it is missing some variables. Thus, removing all data for the
census tract is the easiest thing to do. Thankfully, this only removed
5/2014 census tracts. This method of computing weighted means was later
changed to sum the product of a variable's value and its corresponding
weight, divided by the sum of the weights. In other words, this method
accounted for the population of an area, since census tracts with larger
populations contributed more proportionally to the final conversion from
block groups to census tracts than lower density areas.

The next challenge was getting the exact Chicago boundaries. Initially,
the spatial filtering to visualize Chicago was just using four
coordinates to form a rectangle around the area. The problem with this
was that areas outside the desired analysis were present in the
visualization. Thus, the spatial filtering was later changed to using
places function from the tigris library and transforming the location
into the correct CRS. This filtered the boundaries to the exact polygon.

Then, another persistent challenge to overcome in this analysis was
surprisingly loading in the ChiVes data. Usually, loading in a dataset
should be simple, but the EPA Walkability variables were being
incorporated into the chicago_city final data before the ChiVes
variables had been incorporated into that same variable. This was solved
by performing a left_join() on the EPA Walkability's TRACT_GEOID and
ChiVes' geoid variable.

This analysis is only an introductory step in a hopefully ongoing
process to bring attention to improve walkability across not just
Chicago but around all of the Greater Chicago Metropolitan area as a
whole. Implementing this into ChiVes would vastly increase the chances
of new policy implementation such as new transit lines, added bike
lanes, re-zoning projects, and new street layout projects. As previously
mentioned four paragraphs back, a walkability score does not account for
everything related to the convenience of pedestrians. Future work for
this project could include adding data to integrate these factors into
new visualizations. Perhaps an entirely new data generating process
(DGP) is needed in order to maximize the accuracy of this data. In fact,
the idea that 1090 census tracts are classified as "Above Average
Walkability" does not make sense. If the average walkability score is
singular value, it makes sense why there is no category called "Average
Walkability"; Everything can only either be below or above that average.
Finally, training machine learning models can speed up and also increase
the accuracy of the aforementioned DGP. These models would be able to
predict walkability scores based off of current built-environment
variables, create new variables, and even incorporate scenario-based
testing.

All of these goals and this current analysis is extremely important for
the environmental sustainability and quality of life in the urban form.

------------------------------------------------------------------------

# **References and .csv Export**

```{r export-to-csv}

chicago_boundary <- places(state = "IL", cb = TRUE) %>%
  filter(NAME == "Chicago") %>%
  st_transform(4326)

chicago_export <- final_data %>%
  st_filter(chicago_boundary, .predicate = st_intersects) %>%
  st_drop_geometry() %>%
  select(
    geoid = TRACT_GEOID,
    walkability_score = walk_score,
    walkability_category = walk_category,
    employment_entropy = employment_entropy,
    intersection_density = intersection_density,
    transit_distance_km = transit_distance_km,
  )

write_csv(chicago_export, "chicago_walkability_for_chives.csv")

```

-   U.S. Environmental Protection Agency. (2021). National Walkability
    Index: User Guide and Methodology.
-   U.S. Environmental Protection Agency. (2021). Smart Location
    Database Technical Documentation.
-   ChiVes Project Team. (2024). ChiVes: Chicago Visualization and
    Environmental Sustainability Platform. University of Illinois at
    Urbana-Champaign.

------------------------------------------------------------------------
